Linear stability of magnetized massive protoplanetary disks

### 2014-04-03 paper/linear.tex
tex updates

\subsection{Domain and boundary conditions}

\subsubsection{Case 1: vertical field}%because SG can only destabilize
                                %symmetric density perts
We impose $\bm{U}$ to be an even function of $z$, so that
$d\bm{U}/dz=0$ at $z=0$. This reduces the computational domain to
$z\in[0,Z_s]$ and permits higher numerical resolution. At the disk
surface we set %$\dbx=\dby=\dvz=0$ 
\begin{align}
  \dbx(Z_s) = \dby(Z_s) = \dvz (Z_s) = 0, 
\end{align}
so the field remains vertical. The potential boundary
condition is 
\begin{align}
  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0 \label{pot_bc1},
\end{align}
and is adapted from \cite{goldreich65a}. The general form of this
potential boundary condition is used for case 2 below.

@@

%We adopt two kinds of vertical boundary conditions depending on the
%field geometry. 
%In both case 1 and case 2 the potential boundary
%condition is 
%\begin{align}
%  \dphi^\prime(\zmax) \pm k_x\dphi(\zmax) = 0
%\end{align}

%At the vertical disk boundaries $z=\pm\zmax$ we assume  
%$\dvz=\dbx=\dby=0$, and

%for the potential perturbation \citep[see][]{goldreich65a}, where the
%upper (lower) sign applies to the upper (lower) disk surface. 
%For Case 1 (purely vertical field) we additionally assume $\bm{U}$ is
%an even function of $z$, so that $d\bm{U}/dz=0$ 
%at $z=0$.

%This implies that $\dbx=\dby=0$ at the midplane, consistent
%with a highly resistive dead zone.

\subsection{Numerical procedure}
We use a pseudo-spectral method to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where 
\begin{align}
  \psi_k  = 
  \begin{cases}
    T_{2(k-1)} & B_y \equiv 0 \text{ (case 1)},\\
    T_{k-1}   & B_y\neq 0 \text{ (case 2)},
  \end{cases}
\end{align}
and  $T_l$ is a Chebyshev polynomial of the first kind of order $l$
\citep{stegun65}. Note that for case 1 the mid-plane symmetry condition
is taken care of by the choice of basis functions. 
%Note that in case 1
%The number of basis functions is
%$N_z$ and $l_\mathrm{max}=2(N_Z-1)$ is the highest polynomial order. 
%For a vertical field we only use even Chebyshev polynomials 
%to satisfy the symmetry condition at $z=0$, so in case 1 we only
%consider $z\geq 0$.

The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction. We choose these points as the
extrema of $T_{l_\mathrm{max}}$ plus end points, where
$l_\mathrm{max}$ is the highest polynomial order. Our standard
resolution is $N_z=256$ ($N_z=257$) for case 1 (case 2).

@@

\subsubsection{Non-dimensionalization}\label{non-dim}
We solve the linearized equations in non-dimensional form,
by defining
%We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  \hat{k}_x/H,\quad \sigma = \hat{\sigma}\Omega,
  \quad \delta\bm{v} = \csmid 
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B_z\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi},  
\end{align} 
where $\hat{c}_s=c_s/\csmid$. We also non-dimensionalize other
background quantities, i.e. $\hat{v}_A=v_A/\csmid$,
$\hat{S}=S/\Omega$, $\hat{\kappa}=\kappa/\Omega$,
$\hat{\Omega}_z=\Omega_z/\Omega$ and $\hat{\eta} = \eta/(H^2\Omega)$.

\subsection{Diagnostics}
We visualize results in terms of dimensionless energy densities. We
define 
\begin{align}
  &E_m \equiv \frac{|\delta \hat{\bm{B}}|^2}{2\beta},\\
  &E_g = 
  \frac{\hat{\rho}}{2\hat{c}_s^2}\left|\real\left(\hat{W}\dd\hat{\Phi}^*\right)\right|,\\  
  &E_k = \frac{1}{2}\hat{\rho}|\delta\hat{\bm{v}}|^2,\\
  &E_t = \frac{\hat{\rho}|\hat{W}|^2}{2\hat{c}_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. Although we do not
solve an energy equation, we nevertheless define $E_t$ as a measure of
density perturbations \citep{kojima89}. The total energy
is $E=E_m+E_g+E_k+E_t$. We use $\avg{\cdot}$ to denote an average over
$z$. 
%The colorbar quantifies the importance of
%self-gravity with

@@

%and we then normalize $E_i$ by
%$\mathrm{max}(E)$.

>>>>

\subsection{Domain and boundary conditions}\label{domain}

\subsubsection{Case 1: vertical field}
We impose $\bm{U}$ to be an even function of $z$, so that
$d\bm{U}/dz=0$ at $z=0$. This permits higher numerical resolution by
reducing the computational domain to $z\in~[0,Z_s]$.  
At the upper disk boundary $z=Z_s$ we set 
\begin{align}
  \dbx(Z_s) = \dby(Z_s) = \dvz (Z_s) = 0, 
\end{align}
so the field remains vertical. The potential boundary
condition is 
\begin{align}
  \dphi^\prime(\zmax) + k_x\dphi(\zmax) = 0 \label{pot_bc1},
\end{align}
and is adapted from \cite{goldreich65a}. The general form of this
potential boundary condition is used for case 2 below.

@@

\subsection{Numerical procedure}
We use a pseudo-spectral method to solve the set of linearized
equations. Let
\begin{align}\label{cheby_expand}
  \bm{U}(z) 
  = \sum_{k=1}^{N_z} \bm{U}_k\psi_k(z/\zmax), 
\end{align}
where 
\begin{align}
  \psi_k  = 
  \begin{cases}
    T_{2(k-1)} & B_y \equiv 0 \text{ (case 1)},\\
    T_{k-1}   & B_y\neq 0 \text{ (case 2)},
  \end{cases}
\end{align}
and  $T_l$ is a Chebyshev polynomial of the first kind of order $l$
\citep{stegun65}. Note that for case 1 the midplane symmetry condition
is taken care of by the choice of basis functions.

The pseudo-spectral coefficients $\bm{U}_n$  are obtained by demanding
the set of linear equations to be satisfied at $N_z$ collocation
points along the vertical direction, here chosen to be the 
extrema of $T_{l_\mathrm{max}}$ plus end points, where 
$l_\mathrm{max}$ is the highest polynomial order. Our standard
resolution is $N_z=256$ ($N_z=257$) for case 1 (case 2).

@@

\subsubsection{Non-dimensionalization}\label{non-dim}
We solve the linearized equations in non-dimensional form,
by defining
%We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  \hat{k}_x/H,\quad \sigma = \hat{\sigma}\Omega,
  \quad \delta\bm{v} = \csmid 
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B_z\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi},  
\end{align} 
where $\hat{c}_s=c_s/\csmid$. We also non-dimensionalize 
background quantities, i.e. $\hat{v}_A=v_A/\csmid$,
$\hat{S}=S/\Omega$, $\hat{\kappa}=\kappa/\Omega$,
$\hat{\Omega}_z=\Omega_z/\Omega$ and $\hat{\eta} = \eta/(H^2\Omega)$.

\subsection{Diagnostics}
We visualize results in terms of dimensionless energy densities. We
define 
\begin{align}
  &E_m \equiv \frac{|\delta \hat{\bm{B}}|^2}{2\beta},\\
  &E_g = 
  \frac{\hat{\rho}}{2\hat{c}_s^2}\left|\real\left(\hat{W}\dd\hat{\Phi}^*\right)\right|,\\  
  &E_k = \frac{1}{2}\hat{\rho}|\delta\hat{\bm{v}}|^2,\\
  &E_t = \frac{\hat{\rho}|\hat{W}|^2}{2\hat{c}_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. Although we do not
solve an energy equation, we nevertheless define $E_t$ as a measure of
density perturbations \citep{kojima89}. The total energy
is $E=E_m+E_g+E_k+E_t$. We use $\avg{\cdot}$ to denote an average over
$z$.

@@

### 2014-04-03 paper/paper.tex
tex updates

\section{Introduction}
Astrophysical disks host a wide range of fluid instabilities. Among
them, the magneto-rotational instability \citep[MRI,][]{chandrasekhar61,
  balbus91,balbus98} %balbus %chandrasekhar ref
and gravitational instability \citep[GI, ][]{toomre64,goldreich65a,goldreich65b} 
provide robust pathways to turbulent angular momentum transport that
enables mass accretion \citep[][ 
  and references therein]{balbus99,armitage11,turner14}.   
%gammie, lin/pringle, paczynski 
%relevance of MRI and GI to planet formation 
They are also relevant to planet formation theory. For example,
the strength of MRI turbulence directly affect planetesimal dynamics
in protoplanetary disks \citep{yang12,gressel12}; while GI can
potentially form giant planets directly through disk fragmentation
\citep{boss97,boss98,gammie01,voro13,helled14}. 
%gammie,
%boss %some review articles (PP6)  
%shakura73, pacz78, lin87,

@@

PPDs subject to both MRI and GI are often modeled through separate
turbulent viscosities in a hydrodynamical framework 
\citep{terquem08}. This implicitly assumes that the development of MRI 
and GI can be assessed independently. 
%That is, MRI is unaffected by
%disk self-gravity; and GI is unaffected by weak magnetization. 
Circumstellar disk models that explicitly combine the equations of magneto-hydrodynamics
and self-gravity have been limited to a few early simulations 
\citep{fromang04,fromang04b,fromang05}. It will be neccessary to %no
%major difference % kim/ostriker/stone
revisit and extend these pioneering calculations to fully explore the
impact of MRI and GI on the structure and evolution of PPDs. In
preparation of this, it is important to have a thorough understanding
of the stability properties of such systems.  
%non-ideal effects important

%why is GI and MRI independent 
%compressibility is not fundamental to MRI 
%horizontal wavenumber (gammie/balbus) 
%The effect of self-gravity on the linear MRI has not been considered, 
%since fluid compressibility is not fundamental to the MRI.

Since compressibility is not fundamental for the MRI, much of the
early stability calculations assume incompressible perturbations
\citep{goodman94,jin96}. However, recent works indicate
compressibility may be important under certain conditions,
such as strong fields \citep{kim00, pessah05,bonanno07}.   
Previous MRI studies have also focused on modes with vanishing
radial wavenumber, because they are the most unstable
\citep{sano99,reyes01}. Self-gravity has minimal effect on such
perturbations in a rotating disk. However, modes
with radial lengthscales on the order of the disk scale height may be
subject to self-gravity. It is of interest to generalize the MRI with
non-zero radial waveumbers to massive disks.

%Simiarly, studies of GI in accretion disks have mostly neglected
%magnetization \citep{mamat10,kim12}. 
%particularly in the context of

%Studies of GI in accretion disks do not often include magnetization
%\citep{mamat10,kim12}.

The effect of a magnetic field on the GI of rotating disks has been
considered by some authors.  
% \cite{lynden-bell66} found that a toroidal
%field can destabilize non-axisymmetric disturbances
Recently, \cite{lizano10} generalized the classic Toomre stability
criterion for razor-thin disks to include a vertical field. For
circumstellar disks, the authors concluded that the field is
stabilizing. This is consistent with previous analysis by
\cite{nakamura83} for three-dimensional (3D) uniformly rotating
disks. However, the GI of 3D differentially rotating disks have
mostly neglected magnetic fields \citep{mamat10,kim12}, but such disks
are subject to the MRI if magnetized. 
%how is the dispersion relation for GI affected by the presence of MRI?
%It is interesting to ask whether
%or not the presence of MRI affects GI. 
%
%It is desirable to clarify
%whether or not the presence of MRI modes affect the dispersion
%relation for GI.

%although MRI and GI are physically distinct 
%explicitly show

This paper marks the beginning of our study of magnetized,
self-gravitating PPDs. We start from linear calculations. 
Linear stability analyses have the advantage that a wide range of
parameters can be studied with negligible computational cost. This
allows one to indentify conditions, if any, under which MRI and GI
cannot be consdered independent. It is also important to have such 
calculations at hand to help benchmark and guide non-linear
simulations.

%This paper marks the beginning of our study of magnetized,
%self-gravitating PPDs and is organized as follows.

This paper is organized as follows. \S\ref{setup} lists the governing
equations and describes the disk equilibra under consideration. 
The linear problem is formulated in \S\ref{linear}. The impact
of self-gravity on the MRI is discussed in
\S\ref{result1}; gravitationally unstable disks are considered in
\S\ref{result2} and equilibra with an azimuthal field is explored in
\S\ref{result3}. We summarize results in \S\ref{summary}, including a
discussion of caveats and future directions.

>>>>

\section{Introduction}
Astrophysical disks host a wide range of fluid instabilities. Among
them, the magneto-rotational instability \citep[MRI,][]{chandrasekhar61,
  balbus91,balbus98} 
and gravitational instability \citep[GI, ][]{toomre64,goldreich65a,goldreich65b} 
provide robust pathways to turbulent angular momentum transport that
enables mass accretion \citep[][ 
  and references therein]{balbus99,armitage11,turner14}.   
They are also relevant to planet formation theory. For example,
the strength of MRI turbulence directly affect planetesimal dynamics
in protoplanetary disks \citep{yang12,gressel12}; while GI can
potentially form giant planets directly through disk fragmentation
\citep{boss97,boss98,gammie01,voro13,helled14}.

@@

PPDs subject to both MRI and GI are often modeled through separate
turbulent viscosities in a hydrodynamical framework 
\citep{terquem08}. This implicitly assumes that the development of MRI 
and GI can be assessed independently. 
Circumstellar disk models that explicitly combine the equations of magneto-hydrodynamics
and self-gravity have been limited to a few early simulations 
\citep{fromang04,fromang04b,fromang05}. It will be neccessary to 
revisit and extend these pioneering calculations to fully explore the
impact of MRI and GI on the structure and evolution of PPDs. In
preparation of this, it is important to have a thorough understanding
of the stability properties of such systems.

Since compressibility is not fundamental for the MRI, much of the
early stability calculations assume incompressible perturbations
\citep{goodman94,jin96}. However, recent works indicate
compressibility may be important under certain conditions,
such as strong fields \citep{kim00, pessah05,bonanno07}.   
Previous MRI studies have also focused on modes with vanishing
radial wavenumber, because they are the most unstable
\citep{sano99,reyes01}. Self-gravity has minimal effect on such
perturbations in a rotating disk. However, modes
with radial lengthscales on the order of the disk scale height may be
subject to self-gravity. It is therefore of interest to generalize the MRI with
non-zero radial waveumbers to massive disks.

The effect of a magnetic field on the GI of rotating disks has been
considered recently by \cite{lizano10}, who generalized the Toomre stability
criterion for razor-thin disks to include a vertical field. For
circumstellar disks, the authors concluded that the field is
stabilizing. This is consistent with previous analysis by
\cite{nakamura83} for three-dimensional (3D) uniformly rotating
disks. However, the GI of 3D differentially rotating disks have
mostly neglected magnetic fields \citep{mamat10,kim12}, but such disks
are subject to the MRI if magnetized.

This work marks the beginning of our study of magnetized,
self-gravitating PPDs. We start from linear calculations, which  
have the advantage that a wide range of parameters can be studied at
negligible computational cost. This  allows us to indentify
conditions, if any, under which MRI and GI 
cannot be considered independent. It is also important to have such 
calculations to benchmark and guide future non-linear
simulations.

This paper is organized as follows. \S\ref{setup} lists the governing
equations and describes the disk equilibra under consideration. 
The linear problem is formulated in \S\ref{linear}. The impact
of self-gravity on the MRI with a vertical field is discussed in
\S\ref{result1}, gravitationally unstable disks are considered in
\S\ref{result2}, and equilibra including an azimuthal field is explored in
\S\ref{result3}. We summarize results in \S\ref{summary} with a 
discussion of important extensions to our current models.

### 2014-04-03 paper/results.tex
tex updates

\subsection{Influence of self-gravity on the MRI through the
  background equilibrium}
Here we use polytropic disks, so that the disk thickness is
well-defined. The upper disk boundary is set to 
$Z_s=0.99H$. We fix $\beta=100$ and $k_xH=0.1$ unless otherwise
stated.

%kx -> 0 so incompressible

\subsubsection{Uniform
  resistivity}   
%We calculate the MRI as a function of $Q$ with uniform
%resistivity ($A=1$) for mid-plane Elsasser numbers $\Lambda_0\in[0.3,10]$.    
Fig. \ref{compare_growth_poly_uniresis} plots MRI growth rates as a
function of $Q$ and $\Lambda_0$. The resistivity is uniform ($A=1$). 
%As expected growth rates generally decrease with decreasing 
%$\Lambda_0$ \citep{jin96}. 
In the limit of ideal MHD ($\Lambda_0>1$), there is negligible
dependence on $Q$. In the resistive limit 
($\Lambda_0<1$), however, growth rates decrease noticeably for $Q<0.5$
($Q_\mathrm{2D}\lesssim 1.5$). This shows 
show that disk self-gravity affects the MRI through the background
equilibrium, even though density and potential perturbations are  
unimportant (i.e. the linear response is non-self-gravitating).

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{MRI growth rates as a function of $Q$ for mid-planee 
    Elsasser numbers $\Lambda_0=0.3$ (solid), $0.5$ (dotted), $1.0$
    (dashed) and $5.0$ (dash-dot).  
    \label{compare_growth_poly_uniresis}}
\end{figure}

@@

Since $\hat{\rho}$ is weakly dependent
on $Q$ (Fig. \ref{eqm_den}), self-gravity only affects the
MRI through the factor $f$, which increases with decreasing $Q$ (see
Fig. \ref{plot_fq} in Appendix \ref{appen1}).   
This implies that sufficiently strong self-gravity can stabilize the
MRI by making $ 2H<\lambda $.  %these perts are not gravitationally
                               %unstable

In the ideal limit, we find $\lambda < 2H$ throughout most of the disk
for the values of $Q$ considered, so self-gravity does not affect
growth rates significantly. However, the wavelength of the
instability, in units of $H$, increases. This is shown in
Fig. \ref{compare_result_lambda10} which plots the magnetic energies
for $\Lambda_0=10$ and a range of $Q$ values. The number of vertical
nodes decrease with $Q$, i.e. the disk accommodates fewer wavelengths
because increasing vertical self-gravity makes it thinner.

%Weak variations in growth rates in the ideal limit, seen in 
%Fig. \ref{compare_growth_poly_uniresis}, were found to be associated
%with changes in the mode character: $\gamma$ decreases
%slightly when the mode begins to acquire an additional vertical node as $Q$ is
%increased.

%{\bf DISCUSS STRESS} Although for ideal MHD self-gravity makes no
%difference to the MRI growth rate, we find the associated stress does
%change with $Q$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_lambda10}
  \caption{Magnetic energies as a function of height, in the limit of
    ideal MHD, for various strengths of self-gravity.   
    \label{compare_result_lambda10}}
\end{figure}

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not completely surpressed since
Eq. \ref{sano_crit} is only an approximation.) Although the function
$f(Q)$ does not change significantly for the range of $Q$ considered, the
dependence of $\lambda_\mathrm{resis}$ on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. For resistive cases
we find no nodes in the magnetic energy $E_m$ except at end points,
i.e. only the longest wavelength survives.

@@

\subsubsection{Layered 
  resistivity} 
Here we consider disks with mid-plane Elsasser number $\Lambda_0=0.1$
and a variable resistivity profile with
$A=10^2$. Fig. \ref{poly_layer} compares the magnetic  
energies for $Q=0.2,\,1$ and $4$. They have similar growth rates, $\gamma/\Omega
= 0.53,\,0.64$ and $0.66$, respectively. In the non-self-gravitating
limit ($Q=4$), the MRI is effectively surpressed for
$z\lesssim0.5H$. This is consistent with layered accretion. However,
in the massive disk ($Q=0.2$) the mode occupies a wider vertical
extent because its wavelength (in units of $H$) is larger. This
suggests that in massive disks the MRI is not well localized to the
active layer.

%MRI penetrates 
%d is not as well localized

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_results_poly_layer_amp100}
  \caption{Magnetic energies as a function of height, for layered polytropic disks with
    such that the conductivity increases by a 
    factor of $10^2$ in going from the mid-plane to the upper disk
    boundary. The vertical lines indicate $\Lambda=1$ for each value
    of $Q$.
    \label{poly_layer}}
\end{figure}

\subsubsection{Dependence on $k_x$}
The above experiments show that with increasing 
disk self-gravity, the MRI becomes more global in the vertical
direction. We find a similar result in the horizontal direction. 
In Fig. \ref{compare_growth_poly_kx} we plot MRI growth rates as a
function of $k_x$ for a range of $Q$ values. Increasing self-gravity
decreases the cut-off horizontal wavenumber for the MRI. We checked
that these modes have negligible density perturbations. Then we can 
understand this result by invoking the instability criteria for  
incompressible MRI in an unstratified Keplerian disk,
\begin{align}
  v_A^2(k_z^2 + k_x^2) < 3\Omega^2,
\end{align}
where $k_z$ is a vertical wavenumber \citep{kim00}. Setting $k_z^2\sim
\Omega^2/v_A^2$ and non-dimensionalizing, we find
\begin{align} 
  k_xH \lesssim \frac{\sqrt{\beta}}{f},
\end{align}
where order-unity factors have been dropped. Despite a simplistic
approach, this demonstrates that with increasing self-gravity
(increasing $f$), we expect that MRI modes with small radial
lengthscales to be suppressed.

%, implying that
%small-scale MRI does not operate in massive disks.

@@

%these modes are still non-sg
%use beta = 40 so only a small range of kx need to be simulated to
%demonstrate this effect
%{\bf cut-off $k_xH$ decreases with increasing SG -> small-scale MRI
%  stabilized by self-gravity}

\subsection{Influence of self-gravity on the MRI through the linear
  response}  
Our goal here is to examine whether or not self-gravity can amplify
the density perturbations caused by the MRI. We compute unstable modes
in a massive isothermal disk with $Q=0.2$ (corresponding to
$Q_\mathrm{2D}=0.72$), which is still expected to be marginally  
stable to gravitational instability \citep[][who find
a critical value of  $Q\simeq 0.2$]{mamat10}.   
The upper disk boundary is set to $Z_s=H$.

@@

\subsubsection{Resistive disks}
We repeat the above calculation for resistive disks, but fix
$\beta=100$ and vary the mid-plane Elasser number $\Lambda_0$. Growth
rates are shown in Fig. \ref{gravity_energy_resis}. 
Interestingly, in the highly resistive case
the magnetic and gravitational energies can be comparable. For 
$\Lambda_0=0.1$ and $k_xH\simeq1.3$, $\tau\sim 0.3$ which corresponds
to $\avg{E_g}\sim 0.5\avg{E_m}$.  
Fig. \ref{mri_massive_cowling}, we compare the magnetic
energy of this mode to that computed in in the \emph{Cowling  
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally letting $Q\to\infty$ in Eq. \ref{lin_poisson}).  
The growth rate increases slightly when self-gravity is included in the linear
response, since self-gravity is usually destabilizing. However, the
similarity between $\gamma$ and $E_m$ indicates that the instability
in the self-gravitating calculation is fundamentally the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{
    Growth rates of MRI modes in an isothermal self-gravitating
    disk with $Q=0.2$ ($Q_\mathrm{2D}=0.72$) at fixed $\beta=100$, 
    for a range of mid-plane Elasser numbers. The resistivity is
    uniform. The colorbar measures the importance of self-gravity by $\tau$. 
    \label{gravity_energy_resis}}
\end{figure}

@@

 For weak fields in an ideal disk, the MRI has a small vertical
 wavelength, $\lambda \ll H$. The perturbed mass contained within
 $\sim \lambda$ is small and its potential
 perturbation is unimportant. Furthermore, considering the stratified
 disk, $\lambda\ll H$ imply rapid variations in the density
 perturbation across the disk height, averaging to zero, so the
 magnitude of the associated potential perturbation is small. 
% because the gravitational potential is the integral of the density
% field, i
 Self-gravity plays does not affect the MRI in this regime.

>>>>

\subsection{Influence of self-gravity on the MRI through the
  background equilibrium}
Here we use polytropic disks, which have a well-defined disk
thickness. The upper disk boundary is set to $Z_s=0.99H$. We fix
$\beta=100$ and $k_xH=0.1$ unless otherwise stated.

\subsubsection{Uniform
  resistivity}   
Fig. \ref{compare_growth_poly_uniresis} plots MRI growth rates as a
function of $Q$ and $\Lambda_0$. The resistivity is uniform ($A=1$). 
For ideal MHD ($\Lambda_0>1$), there is negligible dependence on
$Q$. In the resistive limit   
($\Lambda_0<1$), however, growth rates decrease noticeably for $Q<0.5$
($Q_\mathrm{2D}\lesssim 1.5$).  Since we find density and potential
perturbations to be negligible (i.e. the linear response is
non-self-gravitating), this shows that disk self-gravity can affect the
MRI through the background equilibrium.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{MRI growth rates as a function of $Q$ and midplane 
    Elsasser numbers, in polytropic disks with $\beta=100$ and uniform
    resistivity.  
    \label{compare_growth_poly_uniresis}}
\end{figure}

@@

Because $\hat{\rho}$ is weakly dependent
on $Q$ (Fig. \ref{eqm_den}), self-gravity only affects the
MRI through the factor $f$, which increases with decreasing $Q$ (see
Fig. \ref{plot_fq} in Appendix \ref{appen1}).   
This implies that sufficiently strong self-gravity can stabilize the
MRI by making $ 2H<\lambda $.

In the ideal limit, we find $\lambda < 2H$ throughout most of the disk
for the values of $Q$ considered, so self-gravity does not affect
growth rates significantly. However, the ratio $\lambda/2H$ does
increase with stronger self-gravity. Consequently, the wavelength of
the instability, in units of $H$, increases. This is shown in
Fig. \ref{compare_result_lambda10} which plots the magnetic energies
for $\Lambda_0=10$ and a range of $Q$ values. The number of vertical
nodes decrease with $Q$, i.e. the disk accommodates fewer wavelengths
because increasing vertical self-gravity makes it thinner.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_result_lambda10}
  \caption{MRI magnetic energies in ideal polytropic disks
    for different strengths of self-gravity.    
    \label{compare_result_lambda10}}
\end{figure}

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not surpressed since
Eq. \ref{lambda_ideal}---\ref{lambda_resis} is only exact for
unstratified disks.) Although the function 
$f(Q)$ does not change significantly for the range of $Q$ considered,
the dependence of $\lambda_\mathrm{resis}$ on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. Modes in
Fig. \ref{lambda_poly_resis} have no nodes in the magnetic energy
$E_m$ except at $z\simeq0,\,H$, i.e. only the longest wavelength survives
with large resistivity. 
%self-gravity tips it over

@@

\subsubsection{Layered 
  resistivity} 
Here we consider disks with midplane Elsasser number $\Lambda_0=0.1$
and a variable resistivity profile with
$A=10^2$. Fig. \ref{poly_layer} compares the magnetic  
energies for $Q=0.2,\,1$ and $4$. They have similar growth rates, $\gamma/\Omega
= 0.53,\,0.64$ and $0.66$, respectively. In the non-self-gravitating
limit ($Q=4$), the MRI is effectively surpressed for
$z\lesssim0.5H$. This is consistent with the picture of layered
accretion proposed for non-self-gravitating disks \citep{gammie96,fleming03}. 
However,in the massive disk ($Q=0.2$) the mode occupies a wider vertical
extent because its wavelength (in units of $H$) is larger. This
suggests that in massive disks, the MRI is not well localized to a
sub-layer within the height, even when the resisitvity has a layered
structure.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_results_poly_layer_amp100}
  \caption{Magnetic energies as a function of height, for polytropic disks
    in which the conductivity increases by a
    factor $A=10^2$ in going from the midplane to the upper disk
    boundary. The vertical lines indicate $\Lambda=1$ for each value 
    of $Q$.
    \label{poly_layer}}
\end{figure}

\subsubsection{Dependence on $k_x$}
The above experiments show that with increasing 
disk self-gravity, the MRI becomes more global in the vertical
direction. We find a similar result in the horizontal direction. 
Fig. \ref{compare_growth_poly_kx} show MRI growth rates as a
function of $k_x$ for a range of $Q$ values. Increasing self-gravity
decreases the cut-off radial wavenumber for the MRI. We checked
that these modes have negligible density perturbations. Then we can 
understand this result by invoking the instability criteria for  
incompressible MRI in an unstratified Keplerian disk,
\begin{align}
  v_A^2(k_z^2 + k_x^2) < 3\Omega^2,
\end{align}
where $k_z$ is a vertical wavenumber \citep{kim00}. Setting $k_z^2\sim
\Omega^2/v_A^2$ and non-dimensionalizing, we find
\begin{align} 
  k_xH \lesssim \frac{\sqrt{\beta}}{f},
\end{align}
where order-unity factors have been dropped. Despite a simplistic
approach, this demonstrates that with increasing self-gravity
(increasing $f$), we expect MRI modes with small radial
lengthscales to be suppressed.

@@

\subsection{Influence of self-gravity on the MRI through the linear
  response}  
Our goal here is to examine whether or not self-gravity can amplify
the density perturbations associated with the MRI. We compute unstable modes
in a massive isothermal disk with $Q=0.2$ (corresponding to
$Q_\mathrm{2D}=0.72$), which is still expected to be marginally  
stable to gravitational instability \citep[][who find
a critical value of  $Q\simeq 0.2$]{mamat10}.   
The upper disk boundary is set to $Z_s=H$.

@@

\subsubsection{Resistive disks}
We repeat the above calculation for resistive disks, but fix
$\beta=100$ and vary the midplane Elasser number $\Lambda_0$. Growth
rates are shown in Fig. \ref{gravity_energy_resis}. 
Interestingly, in the highly resistive case
the magnetic and gravitational energies can be comparable. For 
$\Lambda_0=0.1$ and $k_xH\simeq1.3$, $\tau\sim 0.3$ which corresponds
to $\avg{E_g}\sim 0.5\avg{E_m}$.  
Fig. \ref{mri_massive_cowling}, we compare the magnetic
energy of this mode to that computed in in the \emph{Cowling  
  approximation}, where the Poisson equation is ignored in the 
linearized equations and the potential perturbation set to zero  
(formally letting $Q\to\infty$ in Eq. \ref{lin_poisson}).  
The growth rate increases slightly when self-gravity is included in the linear
response, since self-gravity is usually destabilizing. However, the
similarity between $\gamma$ and $E_m$ indicates that the instability
in the self-gravitating calculation is fundamentally the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_energy_resis}
  \caption{
    Growth rates of MRI modes in an isothermal self-gravitating
    disk with $Q=0.2$ ($Q_\mathrm{2D}=0.72$) at fixed $\beta=100$, 
    for a range of midplane Elasser numbers. The resistivity is
    uniform. The colorbar measures the importance of self-gravity by $\tau$. 
    \label{gravity_energy_resis}}
\end{figure}

@@

 For weak fields in an ideal disk, the MRI has a small vertical
 wavelength, $\lambda \ll H$. The perturbed mass contained within
 $\sim \lambda$ is small and its potential
 perturbation is unimportant. Furthermore, considering the stratified
 disk, $\lambda\ll H$ imply rapid variations in the density
 perturbation across the disk height, averaging to zero, so the
 magnitude of the associated potential perturbation is small. 
% because the gravitational potential is the integral of the density
% field, i
 Self-gravity does not affect the MRI in this regime.
